Exploring Local Explanations of Nonlinear Models Using Animated Linear Projections

### 2023-05-22 jss/cheem_paper.Rmd
edits through-out

---
documentclass: jss
author:
  - name: Nicholas Spyrison
    orcid: 0000-0002-8417-0212
    affiliation: Monash University
    # use this syntax to add text on several lines
    address: |
      | Dept of Human Centred Computing
      | Faculty of Information Technology
    email: \email{spyrison@gmail.com}
    url: https://nspyrison.netlify.app
  - name: Dianne Cook
    orcid: 0000-0002-3813-7155
    address: |
      | Dept of Econometrics and Business Statistics
    affiliation: Monash University
  - name: Przemyslaw Biecek
    orcid: 0000-0001-8423-1823
    affiliation: |
      | University of Warsaw
      | Warsaw University of Technology
title:
  formatted: "Exploring Local Explanations of Nonlinear Models Using Animated Linear Projections"
  plain:     "Exploring Local Explanations of Nonlinear Models Using Animated Linear Projections"
  # For running headers, if needed
  short:     "Exploring Local Explanations"
abstract: >
    The increased predictive power of machine learning models comes at the cost of increased complexity and loss of interpretability, particularly in comparison to parametric statistical models. This trade-off has led to the emergence of eXplainable AI (XAI) which provides methods, such as local explanations (LEs) and local variable attributions (LVAs), to shed light on how a model use predictors to arrive at a prediction. These provide a point estimate of the linear variable importance in the vicinity of a single observation. However, LVAs tend not to effectively handle association between predictors. To understand how the interaction between predictors affects the variable importance estimate, we can convert LVAs into linear projections and use the radial tour. This is also useful for learning how a model has made a mistake, or the effect of outliers, or clustering of observations. The approach is llustrated with examples from categorical (penguin species, chocolate types) and quantitative (soccer/football salaries, house prices) response models. The methods are implemented in the R package cheem, available on CRAN.
keywords:
  # at least one keyword must be supplied
  formatted: [explainable artificial intelligence, nonlinear model interpretability, visual analytics, local explanations, grand tour, radial tour]
  plain:     [explainable artificial intelligence, nonlinear model interpretability, visual analytics, local explanations, grand tour, radial tour]
preamble: >
  \usepackage{amsmath}
  \usepackage{datetime}
output: rticles::jss_article
bibliography: paper.bib
---

@@

<!-- Introduce explanatory vs predictive modeling -->
There are different reasons and purposes for fitting a model. According to the taxonomies of @breiman_statistical_2001 and @shmueli_explain_2010, it can be useful to group models into two types: explanatory and predictive. Explanatory modeling is used for inferential purposes, while predictive modeling focuses solely on the performance of an objective function. The intended use of the model has important implications for its selection and development. Interpretability is critical in explanatory modeling to draw meaningful inferential conclusions, such as which variables most contribute to a prediction or whether some observations are less well fit. Interpretability becomes more difficult when the model is nonlinear. Nonlinear models occur in statistical models with polynomial or interaction terms between quantitative predictors, and almost all computational models such as random forests, support-vector machines, or neural networks [e.g. @breiman_random_2001; @boser_training_1992; @anderson_introduction_1995].

<!-- nonlinear leading to interpretability issues -->
In linear models interpretation of the importance of variables is relatively straight forward, one adjusts for the covariance of multiple variables when examining the relationship with the response. The interpretation is valid for the full domain of the predictors. In nonlinear models one needs to consider the model in small neighborhoods of the domain to make any assessment of variable importance. Even though this is difficult, it is especially important to interpret model fits as we become more dependent on nonlinear models for routine aspects of life to avoid issues described in @stahl-ethics. Understanding how nonlinear models behave when usage extrapolates outside the domain of predictors, either in sub-spaces where few samples were provided in the training set, or extending outside the domain. It is especially important because nonlinear models can vary wildly and predictions can be dramatically wrong in these areas.

@@

<!-- Data visualization tours -->
For multivariate data visualization, a _tour_ [@asimov_grand_1985; @buja_grand_1986; @lee_state_2021] of linear data projections onto a lower-dimensional space, could be an element of XAI, complementing LVAs. <!--Tours are viewed as an animation over minor changes to the projection basis. Structure in a projection can then be explored visually to see which variables contribute to the formation of that structure. The intuition is similar to watching the shadow of a hidden 3D object change as the object is rotated; watching the shape of the shadow change conveys information about the structure and variables of the object.-->
<!-- tours and models -->
Applying tours to model interpretation is recommended in @wickham_visualizing_2015 primarily to examine the fitted model in the space of the data. @cook_interactive_2007 describe the use of tours for exploring classification boundaries and model diagnostics [@Caragea2008; @lee_pptree_2013; @da_silva_projection_2021].
<!-- Manual tours --> 
There are various types of tours. In a _manual_ or radial tour [@cook_manual_1997; @spyrison_spinifex_2020], the path of linear projections is defined by changing the contribution of a selected variable. We propose to use this to scrutinize the LVAs. <!-- Additional interactive elements in a graphical user interface should allow the user to identify an observation of interest and then explore its local explanation by changing variable contribution with the radial tour. -->
<!-- What-if & counterfactual analysis -->
This approach could be considered to be a counter-factual, what-if analysis, such as _ceteris paribus_ ("other things held constant") profiles [@biecek_ceterisparibus_2020]. <!-- These profiles show how an observation's prediction would change from a marginal change in one explanatory variable, given that other variables are held constant. It ignores correlations of the variables and imagines a case that was not observed. In contrast, our approach is a geometric explanation of the factual; it varies contributions of the variables by rotating the basis, a reorientation of the data object. A constraint in our approach is that the basis must remain orthonormal. When the contribution of one variable decreases, the contributions of others necessarily increase such that there is a complete component in that direction. This also ensures that what is seen is strictly a low-dimensional projection from high-dimensional space and is thus an interpretable visualization.-->

<!-- Paper structure -->
The remainder of this paper is organized as follows. Section \ref{sec:explanations} covers the background of the LEs and the traditional visuals produced. Section \ref{sec:tour} explains the tours and particularly the radial manual tour. Section \ref{sec:cheemviewer} discusses the visual layout in the graphical user interface and how it facilitates analysis, data pre-processing, and package infrastructure. Illustrations are provided in Section \ref{sec:casestudies} for a range of supervised learning tasks with categorical and quantitative response variables. These show how the LVAs can be used to get an overview of the model's use of predictors, and to investigate errors in the model predictions. Section \ref{sec:cheemdiscussion} concludes with a summary of the insights gained. The methods are implemented in the __R__ package __cheem__.

@@

<!-- SHAP -->
We will be focusing on SHAP values in this this paper, but the approach is applicable for any method used to calculate the LVAs. SHAP calculates the variable contributions of one observation by examining the effect of other variables on the predictions. The term "SHAP" refers to @shapley_value_1953's method to evaluate an individual's contribution in cooperative games by assessing this player's performance in the presence or absence of other players. @strumbelj_efficient_2010 introduced SHAP for LEs in machine learning models. Variable importance can depend on the sequence in which variables are entered into the model fitting process, thus for any sequence we get a set of variable contribution values for a single observation. These values will add up to the difference between the fitted value for the observation, and the average fitted value for all observations. Using all possible sequence, or permutation, gives multiple values for each variable, which are averaged to get the SHAP value for an observation. It can be helpful to standardize variables prior to computing SHAP values, if they have been measured on different scales.

The approach is related to partial dependence plots (see for example, chapter 8 of @molnar2022), used to explain the effect of a variable by predicting the response for a range of values on this variable after fixing the value of all other variables to their mean. Though partial dependence plots are a global approximation of the variable importance, while SHAP is specific to one observation.

@@

<!-- Fifa example -->
We use 2020 season FIFA data [@leone_fifa_2020] to illustrate SHAP following the procedures described in @biecek_explanatory_2021. There are 5000 observations of nine predictor variables measuring players' skills and one response variable, wages (in euros). A random forest model is fit regressing players' wages on the skill variables. In this illustration in Figure \ref{fig:shapdistrbd} the SHAP values are compared for a star offensive player (L. Messi) and a prominent defensive player (V. van Dijk). We are interested in knowing how the skill variables locally contribute to the wage prediction of each player. A difference in the attribution of the variable importance across the two positions of the players can be expected. This would be interpreted as how a player's salary depends on which combination of skills. Panel (a) is a version of a breakdown plot [@gosiewska_ibreakdown_2019] where just three sequences of variables are shown, for two observations. A breakdown plot shows the absolute values of the variable attribution for an observation, usually sorted from highest value to the lowest. There is no scale on the horizontal axis here because values are considered relative to each other. Here we can see how the variable contribution can change depending on sequence, relative to both players. (Note that the order of the variables is different in each plot, because they have been sorted by biggest average contribution across both players.) For all sequences, and for both players `reaction` has the strongest contribution, with perhaps more importance for the defensive player. Then it differs by player: for Messi `offense` and `movement` have the strongest contributions, and for van Dijk it is `defense` and `power`, regardless of the variable sequence.

Panel (b) shows the differences in the player's median values (large dots) for 25 such sequences (tick marks). We can see that the wage predictions for the two players come from different combinations of skills sets, as might be expected for players who's value on the team depends on their offensive or defensive prowess. It is also interesting to see from the distribution of values across the different sequence of variables, that there is some multimodality. For example, look at the SHAP values for `reaction` for Messi, and in some sequences reaction has a much lower contribution than others. This suggests that other variables (`offense`, `movement` probably) can substitute for `reaction` in the wage prediction.

@@

There are numerous R packages currently available which provide functions for computing SHAP values, including `fastshap` [@fastshap], `kernelshap` [@kernelshap], `shapr` [@shapr], `shapviz` [@shapviz], `PPtreeregViz` [@PPtreeregViz], `ExplainPrediction` [@ExplainPrediction], `flashlight` [@flashlight] and the package `DALEX` has many resources [@biecek_dalex_2018]. There are many more packages only available through Github, like `treeshap` [@kominsarczyk_treeshap_2021] that is used for this work. @molnar2022 provides good explanations of the different methods and how to apply them to different models.

@@

<!-- Manual tour -->
The manual tour [@cook_manual_1997] defines its path by changing a selected variable's contribution to a basis to allow the variable to contribute more or less to the projection. The requirement constrains the contribution of all other variables that a basis needs to be orthonormal (column correspond to vectors, with unit length, and orthogonal to each other). The manual tour is primarily used to assess the importance of a variable to structure visible in a projection. It also lends itself to pre-computation queued in advance or computed on-the-fly for human-in-the-loop analysis [@karwowski_international_2006].

@@

```{r radialtour, echo=F, fig.align='center', out.width="99%", fig.cap = "The radial tour allows the user to remove a variable from a projection, to examine the importance of this variable to structure in the plot. Here we have a 1D projection of the penguins data displayed as a density plot. The line segments on the bottom correspond to the coefficients of the variables making up the projection. The structure in the plot is bimodality (left), and the importance of the variable \\textsf{bd} is being explored. As this variable contribution is reduced in the plot (middle, right) we can see that the bimodality decreases. Thus \\textsf{bd} is an important variable contributing to the bimodal structure."}
knitr::include_graphics("./figures/radial_tour.png")
```

@@

<!-- Approximations of the spaces, PC1:2 -->
The visualization is composed of the first two principal components of the data (left) and the attribution (middle) spaces. These single 2D projections will not reveal all of the structure of higher-dimensional space, but they are helpful visual summaries. In addition, a plot of the observed against predicted response values is also provided (Figures \ref{fig:classificationcase}b, \ref{fig:regressioncase}a) to help identify observations poorly predicted by the model. <!-- Interactions --> For classification tasks, color indicates the predicted class, and misclassified observations are circled in red. Linked brushing between the plots is provided, and a tabular display of selected points helps to facilitate exploration of the spaces and the model (shown in Figures \ref{fig:regressioncase}d).

While the comparison of these spaces is interesting, the primary purpose of the global view is to enable the selection of particular observations to explore in detail. We have designed it to enable a comparison between an observation that is interesting is some way, perhaps misclassified, or poorly predicted, relative to an observation with similar predictor values but more expected prediction. For brevity, we call the interesting observation the primary investigation (PI) and the other is the comparison investigation (CI). These observations are highlighted as asterisk and $\times$, respectively.

@@

There are two plots in this part of the interface. The first (Figures \ref{fig:classificationcase}e and \ref{fig:regressioncase}e) is a display of the SHAP values for all observations. This will generally give the global view of variables important for the fit as a whole, but it will also highlight observations which have different patterns. The second plot is the radial tour, which for classification is a density plot of a 1D projection (Figure \ref{fig:classificationcase}f), and for regression are scatterplots of the observed response values, and residuals, against a 1D projection (Figure \ref{fig:regressioncase}f).

<!-- Local explanations as par coords -->
The LVAs for all observations are normalized (sum of squares equals 1), and thus, the relative importance of variables can be compared across all observations. These are depicted as a vertical parallel coordinate plot [@ocagne_coordonnees_1885]. (The SHAP values of the PI and CI are shown as dashed and dotted lines, respectively.) One should obtain a sense of the overall importance of variables from this plot. The more important variables will have larger values, and in the case of classification tasks variables which have different magnitude for different classes are more globally important. For example, Figure \ref{fig:classificationcase}e suggests that `bl` is important for distinguishing the green class from the other two. For regression, one might generally observe which variables have low values for all observations (not important). For example, `BMI` and `pwr` in Figure \ref{fig:regressioncase}e, have a range of high and low values (e.g., `off`, `def`), suggesting they are important for some observations and not important for others.

<!-- Segue of PI to attribution projection -->
A bar chart is overlaid to represent the projection shown in the radial tour at right. It starts from the SHAP values of the PI, but if the user changes the projection the length of these bars will reflect this change. (The PI is interactively selected by clicking on a point in the global view). By scaling the SHAP value it becomes an (attribution) projection. <!-- approximates the variable importance for predicting this observation. The combination of variables best explains the difference between the mean response and an observation’s predicted value. It is not an indication of the local shape of the model surface. That is, it is not some indication of the tangent to the curve at this point.-->

@@

```{r regressioncase, echo=F, out.width="100%", fig.cap="Overview of the cheem viewer for regression tasks (quantitative response) and illustration of interactive variables. Panel (a) PCA of the data- and attributions- spaces and the (b) residual plot, predictions by observed values. Four selected points are highlighted in the PC spaces and tabularly displayed. Coloring on a statistic (c) highlights structure organized in the attribution space. Interactive tabular display (d) populates when observations are selected. Contribution of the 1D basis affecting the horizontal position (e) parallel coordinate display of the variable attribution from all observations, and horizontal bars show the contribution to the current basis. Regression projection (f) uses the same horizontal projection and fixes the vertical positions to the observed $y$ and residuals (middle and right)."}
knitr::include_graphics("./figures/app_regression_interactions.PNG")
```

@@

<!-- Exploratory interactions -->
A tooltip displays observation number/name and classification information while the cursor hovers over a point. Linked brushing allows the selection of points (left click and drag) where those points will be highlighted across plots (Figure \ref{fig:classificationcase}a & b). The information corresponding to the selected points is populated on a dynamic table (Figure \ref{fig:classificationcase}d). These interactions aid exploration of the spaces and, finally, the identification of primary and comparison observations.

@@

<!-- Switch to rmd bullets, consistent with thesis -->
- __Data:__ predictors and response are unscaled complete numerical matrix. Most models and local explanations are scale-invariant. Keep normality assumptions of the model in mind. 
- __Model:__ any model and compatible explanation could be explored with this method. Currently, random forest models are applied via the package __randomForest__ [@liaw_classification_2002], compatibility tree SHAP. Modest hyperparameters are used, namely: 125 trees, the number of variables at each split, mtry = $\sqrt{p}$ or $p/3$ for classification and regression, and minimum size of terminal nodes $max(1, n/500)$ or $max(5, n/500)$ for classification and regression.
- __Local explanation:__ Tree SHAP is calculated for _each_ observation using the package __treeshap__ [@kominsarczyk_treeshap_2021]. We opt to find the attribution of each observation in the training data and not fit to fit variable interactions.
- __Cheem viewer:__ after the model and full explanation space are calculated, each variable is scaled by standard deviations away from the mean to achieve common support for visuals. Statistics for mapping to color are computed on the scaled spaces.

<!-- Note on time of execution -->
The time to preprocess the data will vary significantly with the complexity of the model and the LE. For reference, the FIFA data contained 5000 observations of nine explanatory variables took 2.5 seconds to fit a random forest model of modest hyperparameters. Extracting the tree SHAP values of each observation took 270 seconds total. PCA and statistics of the variables and attributions took 2.8 seconds. These run times were from a non-parallelized session on a modern laptop, but suffice to say that most of the time will be spent on the LVA. An increase in model complexity or data dimensionality will quickly become an obstacle. Its reduced computational complexity makes tree SHAP an excellent candidate to start. Alternatively, the package __fastshap__ claims extremely low run times, attributed to fewer calls to the prediction function, partial implementation in C++, and efficient use of logical subsetting, @greenwell_fastshap_2020.

@@

There is more separation visible in the attribution space than in the data space, as would be expected. The predicted vs observed plot reveals a handful of misclassified observations. A Gentoo that has been wrongly labeled as a Chinstrap is selected for illustration. The PI is a misclassified point (represented by the asterisk in the global view and a dashed vertical line in the tour view). The CI is a correctly classified point (represented by an $\times$ and a vertical dotted line).

The radial tour starts from the attribution projection of the misclassified observation (b, left). The important variables identified by SHAP in the (wrong) prediction for this observation are mostly `bl` and `bd` with small contributions of `fl` and `bm`. This projection is a view where the Gentoo (purple) looks much more likely for this observation than Chinstrap. That is, this combination of variables is not particularly useful because the PI looks very much like other Gentoo penguins. The radial tour is used to vary the contribution of flipper length (`fl`) to explore this. (In our exploration, this was the third variable explored. It is typically helpful to explore the variables with more significant contributions, here `bl` and `bd`. Still, when doing this, nothing was revealed about how the PI differed from other Gentoos). On varying `fl` as it contributes increasingly to the projection (b, right), more and more, this penguin looks like a Chinstrap. This suggests that `fl` should be considered an important variable for explaining the (wrong) prediction.

>>>>

---
documentclass: jss
author:
  - name: Nicholas Spyrison
    orcid: 0000-0002-8417-0212
    affiliation: Monash University
    # use this syntax to add text on several lines
    address: |
      | Dept of Human Centred Computing
      | Faculty of Information Technology
    email: \email{spyrison@gmail.com}
    url: https://nspyrison.netlify.app
  - name: Dianne Cook
    orcid: 0000-0002-3813-7155
    address: |
      | Dept of Econometrics and Business Statistics
    affiliation: Monash University
  - name: Przemyslaw Biecek
    orcid: 0000-0001-8423-1823
    affiliation: |
      | University of Warsaw
      | Warsaw University of Technology
title:
  formatted: "Exploring Local Explanations of Nonlinear Models Using Animated Linear Projections"
  plain:     "Exploring Local Explanations of Nonlinear Models Using Animated Linear Projections"
  # For running headers, if needed
  short:     "Exploring Local Explanations"
abstract: >
    The increased predictive power of machine learning models comes at the cost of increased complexity and loss of interpretability, particularly in comparison to parametric statistical models. This trade-off has led to the emergence of eXplainable AI (XAI) which provides methods, such as local explanations (LEs) and local variable attributions (LVAs), to shed light on how a model use predictors to arrive at a prediction. These provide a point estimate of the linear variable importance in the vicinity of a single observation. However, LVAs tend not to effectively handle association between predictors. To understand how the interaction between predictors affects the variable importance estimate, we can convert LVAs into linear projections and use the radial tour. This is also useful for learning how a model has made a mistake, or the effect of outliers, or the clustering of observations. The approach is illustrated with examples from categorical (penguin species, chocolate types) and quantitative (soccer/football salaries, house prices) response models. The methods are implemented in the R package cheem, available on CRAN.
keywords:
  # at least one keyword must be supplied
  formatted: [explainable artificial intelligence, nonlinear model interpretability, visual analytics, local explanations, grand tour, radial tour]
  plain:     [explainable artificial intelligence, nonlinear model interpretability, visual analytics, local explanations, grand tour, radial tour]
preamble: >
  \usepackage{amsmath}
  \usepackage{datetime}
output: rticles::jss_article
bibliography: paper.bib
---

@@

<!-- Introduce explanatory vs predictive modeling -->
There are different reasons and purposes for fitting a model. According to the taxonomies of @breiman_statistical_2001 and @shmueli_explain_2010, it can be useful to group models into two types: explanatory and predictive. Explanatory modeling is used for inferential purposes, while predictive modeling focuses solely on the performance of an objective function. The intended use of the model has important implications for its selection and development. Interpretability is critical in explanatory modeling to draw meaningful inferential conclusions, such as which variables most contribute to a prediction or whether some observations are less well fit. Interpretability becomes more difficult when the model is nonlinear. Nonlinear models occur in statistical models with polynomial or interaction terms between quantitative predictors, and almost all computational models such as random forests, support vector machines, or neural networks [e.g. @breiman_random_2001; @boser_training_1992; @anderson_introduction_1995].

<!-- nonlinear leading to interpretability issues -->
In linear models interpretation of the importance of variables is relatively straightforward, one adjusts for the covariance of multiple variables when examining the relationship with the response. The interpretation is valid for the full domain of the predictors. In nonlinear models, one needs to consider the model in small neighborhoods of the domain to make any assessment of variable importance. Even though this is difficult, it is especially important to interpret model fits as we become more dependent on nonlinear models for routine aspects of life to avoid issues described in @stahl-ethics. Understanding how nonlinear models behave when usage extrapolates outside the domain of predictors, either in sub-spaces where few samples were provided in the training set, or extending outside the domain. It is especially important because nonlinear models can vary wildly and predictions can be dramatically wrong in these areas.

@@

<!-- Data visualization tours -->
For multivariate data visualization, a _tour_ [@asimov_grand_1985; @buja_grand_1986; @lee_state_2021] of linear data projections onto a lower-dimensional space, could be an element of XAI, complementing LVAs. <!--Tours are viewed as an animation over minor changes to the projection basis. Structure in a projection can then be explored visually to see which variables contribute to the formation of that structure. The intuition is similar to watching the shadow of a hidden 3D object change as the object is rotated; watching the shape of the shadow change conveys information about the structure and variables of the object.-->
<!-- tours and models -->
Applying tours to model interpretation is recommended by @wickham_visualizing_2015 primarily to examine the fitted model in the space of the data. @cook_interactive_2007 describe the use of tours for exploring classification boundaries and model diagnostics [@Caragea2008; @lee_pptree_2013; @da_silva_projection_2021].
<!-- Manual tours --> 
There are various types of tours. In a _manual_ or radial tour [@cook_manual_1997; @spyrison_spinifex_2020], the path of linear projections is defined by changing the contribution of a selected variable. We propose to use this to scrutinize the LVAs. <!-- Additional interactive elements in a graphical user interface should allow the user to identify an observation of interest and then explore its local explanation by changing variable contribution with the radial tour. -->
<!-- What-if & counterfactual analysis -->
This approach could be considered to be a counter-factual, what-if analysis, such as _ceteris paribus_ ("other things held constant") profiles [@biecek_ceterisparibus_2020]. <!-- These profiles show how an observation's prediction would change from a marginal change in one explanatory variable, given that other variables are held constant. It ignores correlations of the variables and imagines a case that was not observed. In contrast, our approach is a geometric explanation of the factual; it varies contributions of the variables by rotating the basis, a reorientation of the data object. A constraint in our approach is that the basis must remain orthonormal. When the contribution of one variable decreases, the contributions of others necessarily increase such that there is a complete component in that direction. This also ensures that what is seen is strictly a low-dimensional projection from high-dimensional space and is thus an interpretable visualization.-->

<!-- Paper structure -->
The remainder of this paper is organized as follows. Section \ref{sec:explanations} covers the background of the LEs and the traditional visuals produced. Section \ref{sec:tour} explains the tours and particularly the radial manual tour. Section \ref{sec:cheemviewer} discusses the visual layout in the graphical user interface and how it facilitates analysis, data pre-processing, and package infrastructure. Illustrations are provided in Section \ref{sec:casestudies} for a range of supervised learning tasks with categorical and quantitative response variables. These show how the LVAs can be used to get an overview of the model's use of predictors and to investigate errors in the model predictions. Section \ref{sec:cheemdiscussion} concludes with a summary of the insights gained. The methods are implemented in the __R__ package __cheem__.

@@

<!-- SHAP -->
We will be focusing on SHAP values in this paper, but the approach is applicable to any method used to calculate the LVAs. SHAP calculates the variable contributions of one observation by examining the effect of other variables on the predictions. The term "SHAP" refers to @shapley_value_1953's method to evaluate an individual's contribution in cooperative games by assessing this player's performance in the presence or absence of other players. @strumbelj_efficient_2010 introduced SHAP for LEs in machine learning models. Variable importance can depend on the sequence in which variables are entered into the model fitting process, thus for any sequence we get a set of variable contribution values for a single observation. These values will add up to the difference between the fitted value for the observation, and the average fitted value for all observations. Using all possible sequences, or permutations, gives multiple values for each variable, which are averaged to get the SHAP value for an observation. It can be helpful to standardize variables prior to computing SHAP values if they have been measured on different scales.

The approach is related to partial dependence plots (for example see chapter 8 of @molnar2022), used to explain the effect of a variable by predicting the response for a range of values on this variable after fixing the value of all other variables to their mean. Though partial dependence plots are a global approximation of the variable importance, while SHAP is specific to one observation.

@@

<!-- Fifa example -->
We use 2020 season FIFA data [@leone_fifa_2020] to illustrate SHAP following the procedures described in @biecek_explanatory_2021. There are 5000 observations of nine predictor variables measuring players' skills and one response variable, wages (in euros). A random forest model is fit regressing players' wages on the skill variables. In this illustration in Figure \ref{fig:shapdistrbd} the SHAP values are compared for a star offensive player (L. Messi) and a prominent defensive player (V. van Dijk). We are interested in knowing how the skill variables locally contribute to the wage prediction of each player. A difference in the attribution of the variable importance across the two positions of the players can be expected. This would be interpreted as how a player's salary depends on which combination of skills. Panel (a) is a version of a breakdown plot [@gosiewska_ibreakdown_2019] where just three sequences of variables are shown, for two observations. A breakdown plot shows the absolute values of the variable attribution for an observation, usually sorted from the highest value to the lowest. There is no scale on the horizontal axis here because values are considered relative to each other. Here we can see how the variable contribution can change depending on sequence, relative to both players. (Note that the order of the variables is different in each plot because they have been sorted by the biggest average contribution across both players.) For all sequences, and for both players `reaction` has the strongest contribution, with perhaps more importance for the defensive player. Then it differs by player: for Messi `offense` and `movement` have the strongest contributions, and for van Dijk it is `defense` and `power`, regardless of the variable sequence.

Panel (b) shows the differences in the player's median values (large dots) for 25 such sequences (tick marks). We can see that the wage predictions for the two players come from different combinations of skill sets, as might be expected for players whose value on the team depends on their offensive or defensive prowess. It is also interesting to see from the distribution of values across the different sequences of variables, that there is some multimodality. For example, look at the SHAP values for `reaction` for Messi, and in some sequences, reaction has a much lower contribution than others. This suggests that other variables (`offense`, `movement` probably) can substitute for `reaction` in the wage prediction.

@@

There are numerous R packages currently available that provide functions for computing SHAP values, including `fastshap` [@fastshap], `kernelshap` [@kernelshap], `shapr` [@shapr], `shapviz` [@shapviz], `PPtreeregViz` [@PPtreeregViz], `ExplainPrediction` [@ExplainPrediction], `flashlight` [@flashlight], and the package `DALEX` has many resources [@biecek_dalex_2018]. There are many more packages only available through Github, like `treeshap` [@kominsarczyk_treeshap_2021] that is used for this work. @molnar2022 provides good explanations of the different methods and how to apply them to different models.

@@

<!-- Manual tour -->
The manual tour [@cook_manual_1997] defines its path by changing a selected variable's contribution to a basis to allow the variable to contribute more or less to the projection. The requirement constrains the contribution of all other variables that a basis needs to be orthonormal (columns correspond to vectors, with unit length, and orthogonal to each other). The manual tour is primarily used to assess the importance of a variable to the structure visible in a projection. It also lends itself to pre-computation queued in advance or computed on the fly for human-in-the-loop analysis [@karwowski_international_2006].

@@

```{r radialtour, echo=F, fig.align='center', out.width="99%", fig.cap = "The radial tour allows the user to remove a variable from a projection, to examine the importance of this variable to the structure in the plot. Here we have a 1D projection of the penguins data displayed as a density plot. The line segments on the bottom correspond to the coefficients of the variables making up the projection. The structure in the plot is bimodality (left), and the importance of the variable \\textsf{bd} is being explored. As this variable contribution is reduced in the plot (middle, right) we can see that the bimodality decreases. Thus \\textsf{bd} is an important variable contributing to the bimodal structure."}
knitr::include_graphics("./figures/radial_tour.png")
```

@@

<!-- Approximations of the spaces, PC1:2 -->
The visualization is composed of the first two principal components of the data (left) and the attribution (middle) spaces. These single 2D projections will not reveal all of the structure of higher-dimensional space, but they are helpful visual summaries. In addition, a plot of the observed against predicted response values is also provided (Figures \ref{fig:classificationcase}b, \ref{fig:regressioncase}a) to help identify observations poorly predicted by the model. <!-- Interactions --> For classification tasks, color indicates the predicted class and misclassified observations are circled in red. Linked brushing between the plots is provided, and a tabular display of selected points helps to facilitate the exploration of the spaces and the model (shown in Figures \ref{fig:regressioncase}d).

While the comparison of these spaces is interesting, the primary purpose of the global view is to enable the selection of particular observations to explore in detail. We have designed it to enable a comparison between an observation that is interesting in some way, perhaps misclassified, or poorly predicted, relative to an observation with similar predictor values but a more expected prediction. For brevity, we call the interesting observation the primary investigation (PI), and the other is the comparison investigation (CI). These observations are highlighted as an asterisk and $\times$, respectively.

@@

There are two plots in this part of the interface. The first (Figures \ref{fig:classificationcase}e and \ref{fig:regressioncase}e) is a display of the SHAP values for all observations. This will generally give the global view of variables important for the fit as a whole, but it will also highlight observations that have different patterns. The second plot is the radial tour, which for classification is a density plot of a 1D projection (Figure \ref{fig:classificationcase}f), and for regression are scatterplots of the observed response values, and residuals, against a 1D projection (Figure \ref{fig:regressioncase}f).

<!-- Local explanations as par coords -->
The LVAs for all observations are normalized (sum of squares equals 1), and thus, the relative importance of variables can be compared across all observations. These are depicted as a vertical parallel coordinate plot [@ocagne_coordonnees_1885]. (The SHAP values of the PI and CI are shown as dashed and dotted lines, respectively.) One should obtain a sense of the overall importance of variables from this plot. The more important variables will have larger values, and in the case of classification tasks variables that have different magnitudes for different classes are more globally important. For example, Figure \ref{fig:classificationcase}e suggests that `bl` is important for distinguishing the green class from the other two. For regression, one might generally observe which variables have low values for all observations (not important). For example, `BMI` and `pwr` in Figure \ref{fig:regressioncase}e, have a range of high and low values (e.g., `off`, `def`), suggesting they are important for some observations and not important for others.

<!-- Segue of PI to attribution projection -->
A bar chart is overlaid to represent the projection shown in the radial tour on the right. It starts from the SHAP values of the PI, but if the user changes the projection the length of these bars will reflect this change. (The PI is interactively selected by clicking on a point in the global view). By scaling the SHAP value it becomes an (attribution) projection. <!-- approximates the variable importance for predicting this observation. The combination of variables best explains the difference between the mean response and an observation’s predicted value. It is not an indication of the local shape of the model surface. That is, it is not some indication of the tangent to the curve at this point.-->

@@

```{r regressioncase, echo=F, out.width="100%", fig.cap="Overview of the cheem viewer for regression tasks (quantitative response) and illustration of interactive variables. Panel (a) PCA of the data- and attributions- spaces and the (b) residual plot, predictions by observed values. Four selected points are highlighted in the PC spaces and tabularly displayed. Coloring on a statistic (c) highlights the structure organized in the attribution space. Interactive tabular display (d) populates when observations are selected. Contribution of the 1D basis affecting the horizontal position (e) parallel coordinate display of the variable attribution from all observations, and horizontal bars show the contribution to the current basis. Regression projection (f) uses the same horizontal projection and fixes the vertical positions to the observed $y$ and residuals (middle and right)."}
knitr::include_graphics("./figures/app_regression_interactions.PNG")
```

@@

<!-- Exploratory interactions -->
A tooltip displays the observation number/name and classification information while the cursor hovers over a point. Linked brushing allows the selection of points (left click and drag) where those points will be highlighted across plots (Figure \ref{fig:classificationcase}a & b). The information corresponding to the selected points is populated on a dynamic table (Figure \ref{fig:classificationcase}d). These interactions aid the exploration of the spaces and, finally, the identification of primary and comparison observations.

@@

<!-- Switch to rmd bullets, consistent with thesis -->
- __Data:__ predictors and response are unscaled complete numerical matrix. Most models and local explanations are scale-invariant. Keep the normality assumptions of the model in mind.
- __Model:__ any model and compatible explanation could be explored with this method. Currently, random forest models are applied via the package __randomForest__ [@liaw_classification_2002], compatibility tree SHAP. Modest hyperparameters are used, namely: 125 trees, the number of variables at each split, mtry = $\sqrt{p}$ or $p/3$ for classification and regression, and minimum size of terminal nodes $max(1, n/500)$ or $max(5, n/500)$ for classification and regression.
- __Local explanation:__ Tree SHAP is calculated for _each_ observation using the package __treeshap__ [@kominsarczyk_treeshap_2021]. We opt to find the attribution of each observation in the training data and not fit to fit variable interactions.
- __Cheem viewer:__ after the model and full explanation space are calculated, each variable is scaled by standard deviations away from the mean to achieve common support for visuals. Statistics for mapping to color are computed on the scaled spaces.

<!-- Note on time of execution -->
The time to preprocess the data will vary significantly with the complexity of the model and the LE. For reference, the FIFA data contained 5000 observations of nine explanatory variables that took 2.5 seconds to fit a random forest model of modest hyperparameters. Extracting the tree SHAP values of each observation took 270 seconds in total. PCA and statistics of the variables and attributions took 2.8 seconds. These run times were from a non-parallelized session on a modern laptop, but suffice it to say that most of the time will be spent on the LVA. An increase in model complexity or data dimensionality will quickly become an obstacle. Its reduced computational complexity makes tree SHAP an excellent candidate to start. Alternatively, some package and methods use approximate calculations of LEs, such as __fastshap__ @greenwell_fastshap_2020.

@@

There is more separation visible in the attribution space than in the data space, as would be expected. The predicted vs observed plot reveals a handful of misclassified observations. A Gentoo which has been wrongly labeled as a Chinstrap is selected for illustration. The PI is a misclassified point (represented by the asterisk in the global view and a dashed vertical line in the tour view). The CI is a correctly classified point (represented by an $\times$ and a vertical dotted line).

The radial tour starts from the attribution projection of the misclassified observation (b, left). The important variables identified by SHAP in the (wrong) prediction for this observation are mostly `bl` and `bd` with small contributions of `fl` and `bm`. This projection is a view where the Gentoo (purple) looks much more likely for this observation than Chinstrap. That is, this combination of variables is not particularly useful because the PI looks very much like other Gentoo penguins. The radial tour is used to vary the contribution of flipper length (`fl`) to explore this. (In our exploration, this was the third variable explored. It is typically helpful to explore the variables with more significant contributions, here `bl` and `bd`. Still, when doing this, nothing was revealed about how the PI differed from other Gentoos). On varying `fl`, as it contributes increasingly to the projection (b, right), more and more, this penguin looks like a Chinstrap. This suggests that `fl` should be considered an important variable for explaining the (wrong) prediction.
